Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2014-05-27 paper.tex
added more to responses. Getting there.

Aside from this extension of Mackenthun's algorithm, the primary purpose of this paper is to analyse the least squares estimators of carrier phase and amplitude obtained by minimising~\eqref{eq:LSdefn}.  In the literature, analytic study of non-data-aided estimators of carrier phase has primarily involved the derivation of Cram\'{e}r-Rao lower bounds under the assumption that the noise $\{w_i\}$ is complex Gaussian distributed~\cite{Delmas_exact_crb_psk_2008,950345,Steendam_lowsnr_crb_2001,Cowley_crbs_phase_freq_1996,DAndrea_modified_bounds_1994}.  The bounds themselves are often complicated to compute and analyse and for this reason it is common to make approximations, known as \emph{modified} Cram\'{e}r-Rao bounds~\cite{DAndrea_modified_bounds_1994,Pollok_mckilliam_crb_cpm_2013}, or to analyse bounds that are valid asymptotically as the noise variance increases to infinity or decreases to zero~\cite{Moeneclaey_limit_crbs_1998,Steendam_lowsnr_crb_2001,Delmas_exact_crb_psk_2008}.  %To the authors' knowledge, all of the existing bounds are applicable only in the fully non-data-aided case or in the completely data-aided case, that is, when either no pilot symbols exist, or when no data symbols exist.  Deriving bounds for the common practical scenario where there is a mixture of pilot and data symbols might be possible, but would further complicate already complicated bounds.

@@

One criticism that might be directed at our asymptotic approach when compared to the Cram\'{e}r-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convincing response is that, while the specific results given here apply only to the least squares estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection}~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010}, would potentially allow this.  
More ambitiously, our techniques might allow analysis of non-data-aided channel estimators for modern communications systems such as mulitple antenna systems, systems employing orthogonal frequency division multiplexing, and systems that must withstand rapidly time varying channels~\cite{4138046,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cram\'{e}r-Rao lower bounds.

@@

We will have particular use of complex random variables of the form $1 + W$ where $W$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + W.
\]
The joint pdf of $R$ and $\Phi$ can be shown to be
 \begin{equation}\label{eq:pdfRPhi}
 f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{2\pi\sqrt{r^2 - 2r\cos\phi + 1}}.
 \end{equation}
Since $\cos\phi$ has period $2\pi$ and is even on $[-\pi,\pi]$ it follows that $f(r,\phi)$ has period $2\pi$ and is even on $[-\pi,\pi]$ with respect to $\phi$.  The mean of $R e^{j\Phi}$ is equal to one because the mean of $W$ is zero.  So,
%\begin{equation}\label{eq:expectRepartRphi}
\[
\expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
\]
%\end{equation}
where $\Re(\cdot)$ denotes the real part, and
\begin{equation}\label{eq:expectImpartRphi}
\expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
\end{equation}
where $\Im(\cdot)$ denotes the imaginary part.

>>>>

Aside from this extension of Mackenthun's algorithm, the primary purpose of this paper is to analyse the least squares estimators of carrier phase and amplitude obtained by minimising~\eqref{eq:LSdefn}.  In the literature, analytic study of non-data-aided estimators of carrier phase has primarily involved the derivation of Cram\'{e}r-Rao lower bounds under the assumption that the noise $\{w_i\}$ is complex Gaussian distributed~\cite{Delmas_exact_crb_psk_2008,950345,Steendam_lowsnr_crb_2001,Cowley_crbs_phase_freq_1996,DAndrea_modified_bounds_1994}.  The bounds themselves are often complicated to compute and analyse and for this reason it is common to make approximations, known as \emph{modified} Cram\'{e}r-Rao bounds~\cite{DAndrea_modified_bounds_1994,Pollok_mckilliam_crb_cpm_2013}, or to analyse bounds that are valid asymptotically as the noise variance increases to infinity or decreases to zero~\cite{Moeneclaey_limit_crbs_1998,Steendam_lowsnr_crb_2001,Noels_crbs_with_pilots_2005,Delmas_exact_crb_psk_2008}.  %To the authors' knowledge, all of the existing bounds are applicable only in the fully non-data-aided case or in the completely data-aided case, that is, when either no pilot symbols exist, or when no data symbols exist.  Deriving bounds for the common practical scenario where there is a mixture of pilot and data symbols might be possible, but would further complicate already complicated bounds.

@@

One criticism that might be directed at our asymptotic approach when compared to the Cram\'{e}r-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convincing response is that, while the specific results given here apply only to the least squares estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection}~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010,McKilliam_pps_unwrapping_tsp_2014}, would potentially allow this.  
More ambitiously, our techniques might allow analysis of non-data-aided channel estimators for modern communications systems such as mulitple antenna systems, systems employing orthogonal frequency division multiplexing, and systems that must withstand rapidly time varying channels~\cite{4138046,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cram\'{e}r-Rao lower bounds.

@@

We will have particular use of complex random variables of the form $1 + W$ where $W$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + W.
\]
The joint pdf of $R$ and $\Phi$ can be shown to be
 \begin{equation}\label{eq:pdfRPhi}
 f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{2\pi\sqrt{r^2 - 2r\cos\phi + 1}}.
 \end{equation}
Since $\cos\phi$ has period $2\pi$ and is even on $[-\pi,\pi]$ it follows that $f(r,\phi)$ has period $2\pi$ and is even on $[-\pi,\pi]$ with respect to $\phi$.  The mean of $R e^{j\Phi}$ is equal to one because the mean of $W$ is zero.  So,
%\begin{equation}\label{eq:expectRepartRphi}
\[
\expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
\]
%\end{equation}
where $\Re(\cdot)$ denotes the real part, and
\begin{equation}\label{eq:expectImpartRphi}
\expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
\end{equation}
where $\Im(\cdot)$ denotes the imaginary part, and $\expect$ denotes the expected value.  Whenever there is no chance of ambiguity we write simply $\expect X$ rather than $\expect(X)$ to denote the expected value of the random variable $X$.  For two random variables $X$ and $Y$ the notation $\expect X Y$ will always mean the expected value of the product $XY$, that is, $\expect(XY)$ and not $\expect(X)Y$.

### 2014-05-27 response2.tex
added more to responses. Getting there.

\newcommand{\fracpart}[1]{\left\langle #1 \right\rangle}
\newcommand{\sfracpart}[1]{\langle #1 \rangle}
\newcommand{\round}[1]{\lfloor #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\ceil}[1]{\lceil #1 \rceil}
\newcommand{\abs}[1]{{\left\vert #1 \right\vert}}
\newcommand{\sabs}[1]{{\vert #1 \vert}}
\newcommand{\reals}{{\mathbb R}}
\newcommand{\ints}{{\mathbb Z}}
\newcommand{\integers}{{\mathbb Z}}
\newcommand{\uy}{\underline{\bm{Y}}}
\newcommand{\uey}{\underline{Y}}
\newcommand{\sign}[1]{\mathtt{sign}(#1)}
\newcommand{\Qbf}{{\mathbf Q}}
\newcommand{\Bbf}{{\mathbf B}}
\newcommand{\Ibf}{{\mathbf I}}
\newcommand{\ybf}{{\mathbf y}}
\newcommand{\xbf}{{\mathbf x}}
\newcommand{\zbf}{{\mathbf z}}
\newcommand{\ebf}{{\mathbf e}}
\newcommand{\vbf}{{\mathbf v}}
\newcommand{\wbf}{{\mathbf w}}
\newcommand{\kbf}{{\mathbf k}}
\newcommand{\sbf}{{\mathbf s}}
\newcommand{\tbf}{{\mathbf t}}
\newcommand{\pbf}{{\mathbf p}}
\newcommand{\Pibf}{{\mathbf \Pi}}
\newcommand{\onebf}{{\mathbf 1}}
\newcommand{\fbf}{{\mathbf f}}
\newcommand{\ubf}{{\mathbf u}}

@@

\item \textbf{Comment:} p. 3: ``all of the exiting bounds are applicable only in the fully non-data-aided case or in the completely data aided case \dots'' --> bounds for the common scenario where there is a mixture of pilot and data symbols can be found in "N. Noels, H. Steendam, M. Moeneclaey en H. Bruneel, Carrier Phase and Frequency Estimation for Pilot Symbol Assisted Transmission: Bounds and Algorithms, IEEE TSP, Vol. 53, No. 12, p. 4578-4587, Dec. 2005. \\
\textbf{Response:} BLERG. I've removed this paragraph from the introduction but will have to update the code and simulations.

@@

\item \textbf{Comment:} p. 5 (and further throughout the text): ``pilots symbols" (with `s') and ``pilot symbols'' (without `s') are both used $\to$ use only 1, I prefer ``pilot symbols'' \\
\textbf{Response:} All occurences of pilots symbols have been changed to pilot symbols.

\item \label{rev2:commentnotationsection} \textbf{Comment:} I think it would be a good idea to specify the main notational conventions at the end of the introductory section \\
\textbf{Response:} It is the authors preference to introduce notation as it is used in the text rather than in a dedicated notation section. %BLERG. This need to be done. It's connected with comment~\ref{rev2:commentE}.

@@

\item \label{rev2:commentE} \textbf{Comment:} I would prefer to see brackets around the functions of which the mean is computed (use of the operator $\mathbb E$) \\
\textbf{Response:} BLERG.  Connected with comment~\ref{rev2:commentnotationsection}.

@@

\item \textbf{Comment:}  although the most important contribution of the paper is with respect to the biased amplitude estimate, only 1 figure is presented for amplitude estimation, namely for $L=2048$ and BPSK $\to$ this seems not well balanced  \\
\textbf{Response:} Figures have been added showing the amplitude estimation error for BSPK, QPSK, and 8-PSK.

@@

\item \textbf{Comment:}  There are many formulas (starting from page 11) with the notation $\fracpart{x}$ meaning ``$x$ taken modulo $2\pi/M$''. These angle brackets don't strike the eye and can be easily confused with ordinary parentheses. To make them much more visible, the authors could use the same approach as for ``$x$ taken modulo $2\pi$'' - by using corresponding subscript, i.e. use $\fracpart{x}_{2\pi/M}$ instead of $\fracpart{x}$. \\
\textbf{Response:}

>>>>

\newcommand{\fracpart}[1]{\left\langle #1 \right\rangle}
\newcommand{\sfracpart}[1]{\langle #1 \rangle}
\newcommand{\round}[1]{\lfloor #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\ceil}[1]{\lceil #1 \rceil}
\newcommand{\abs}[1]{{\left\vert #1 \right\vert}}
\newcommand{\sabs}[1]{{\vert #1 \vert}}
\newcommand{\reals}{{\mathbb R}}
\newcommand{\ints}{{\mathbb Z}}
\newcommand{\integers}{{\mathbb Z}}
\newcommand{\uy}{\underline{\bm{Y}}}
\newcommand{\uey}{\underline{Y}} 
\newcommand{\sign}[1]{\mathtt{sign}(#1)}
\newcommand{\Qbf}{{\mathbf Q}}
\newcommand{\Bbf}{{\mathbf B}}
\newcommand{\Ibf}{{\mathbf I}}
\newcommand{\ybf}{{\mathbf y}}
\newcommand{\xbf}{{\mathbf x}}
\newcommand{\zbf}{{\mathbf z}}
\newcommand{\ebf}{{\mathbf e}}
\newcommand{\vbf}{{\mathbf v}}
\newcommand{\wbf}{{\mathbf w}}
\newcommand{\kbf}{{\mathbf k}}
\newcommand{\sbf}{{\mathbf s}}
\newcommand{\tbf}{{\mathbf t}}
\newcommand{\pbf}{{\mathbf p}}
\newcommand{\Pibf}{{\mathbf \Pi}}
\newcommand{\onebf}{{\mathbf 1}}
\newcommand{\fbf}{{\mathbf f}}
\newcommand{\ubf}{{\mathbf u}}
\newcommand{\expect}{{\mathbb E}}

@@

\item \textbf{Comment:} p. 3: ``all of the exiting bounds are applicable only in the fully non-data-aided case or in the completely data aided case \dots'' --> bounds for the common scenario where there is a mixture of pilot and data symbols can be found in "N. Noels, H. Steendam, M. Moeneclaey en H. Bruneel, Carrier Phase and Frequency Estimation for Pilot Symbol Assisted Transmission: Bounds and Algorithms, IEEE TSP, Vol. 53, No. 12, p. 4578-4587, Dec. 2005. \\
\textbf{Response:} We have added this reference and have removed the sentence begining ``To the authors' knowledge, all of the existing bounds are applicable only in the fully non-data-aided case \dots''. Note that the model considered in the paper by Noels~et.~al. treats the amplitude as fixed and known and includes a frequency offset term.  Thus, the bounds are not immediately applicable in our setting and we are not able to meaningfully compare these bounds with our simulation results.  For this reason with have kept with the bound described in~\cite{950345} as in the previous version of the manuscript.

@@

\item \textbf{Comment:} p. 5 (and further throughout the text): ``pilots symbols" (with `s') and ``pilot symbols'' (without `s') are both used $\to$ use only 1, I prefer ``pilot symbols'' \\
\textbf{Response:} All occurences of ``pilots symbols'' have been changed to ``pilot symbols''.

\item \label{rev2:commentnotationsection} \textbf{Comment:} I think it would be a good idea to specify the main notational conventions at the end of the introductory section \\
\textbf{Response:} We request not to make this change.  It is the authors' preference to introduce notation as it is used in the text rather than in a dedicated notation section. %BLERG. This need to be done. It's connected with comment~\ref{rev2:commentE}.

@@

\item \label{rev2:commentE} \textbf{Comment:} I would prefer to see brackets around the functions of which the mean is computed (use of the operator $\mathbb E$) \\
\textbf{Response:} We prefer not to use brackets whenever there is no chance of ambiguity.  To clarify this for the reader the following sentences have been added to Section~\ref{sec:stat-prop-least}:

``Whenever there is no chance of ambiguity we write simply $\expect X$ rather than $\expect(X)$ to denote the expected value of the random variable $X$.  For two random variables $X$ and $Y$ the notation $\expect X Y$ will always mean the expected value of the product $XY$, that is, $\expect(XY)$ and not $\expect(X)Y$.''

The bracketless notation is heavily employed in the statistical literature on empirical processes~\cite{Pollard_new_ways_clts_1986,Pollard_conv_stat_proc_1984,Pollard_asymp_empi_proc_1989,van2009empirical}. The notation would become heavily cluttered with brackets otherwise.  Our proofs in the appendix rely upon this literature.

@@

\item \textbf{Comment:}  although the most important contribution of the paper is with respect to the biased amplitude estimate, only 1 figure is presented for amplitude estimation, namely for $L=2048$ and BPSK $\to$ this seems not well balanced  \\
\textbf{Response:} Figures have been added showing the amplitude estimation error for BSPK, QPSK, and 8-PSK. BLERG: Add to text!

@@

\item \textbf{Comment:}  There are many formulas (starting from page 11) with the notation $\fracpart{x}$ meaning ``$x$ taken modulo $2\pi/M$''. These angle brackets don't strike the eye and can be easily confused with ordinary parentheses. To make them much more visible, the authors could use the same approach as for ``$x$ taken modulo $2\pi$'' - by using corresponding subscript, i.e. use $\fracpart{x}_{2\pi/M}$ instead of $\fracpart{x}$. \\
\textbf{Response:} We request not to make this change.  There are two reasons.  The first is that this notation leads to situations where there are subscripts withing superscripts.  For example, in the proof of Theorem 1 we would have expressions of the form
\[
\rho_0 R_i e^{j\sfracpart{\lambda + \Phi_i}_{2\pi/M}}
\]
and in the appendix we would obtain expressions such as
\[
2 \hat{\rho} \rho_0 R_k \Re\left( \eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}_{2\pi/M}}\right)
\]
and
\[
q_2(\lambda) + j k_2(\lambda) = \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}_{2\pi/M}} - R_1 e^{j\fracpart{\Phi_1}_{2\pi/M}}  \big].
\]
We feel that these are more difficult to parse that the existing simpler
\[
\rho_0 R_i e^{j\sfracpart{\lambda + \Phi_i}},
\]
\[
2 \hat{\rho} \rho_0 R_k \Re\left( \eta e^{j\sfracpart{\hat{\lambda}_L + \Phi_k}}\right),
\]
and
\[
q_2(\lambda) + j k_2(\lambda) = \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}} - R_1 e^{j\fracpart{\Phi_1}}  \big].
\]

The second reason is that some of the authors have made use of the notation $\fracpart{\cdot}$ is previous work~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010,McKilliam_pps_unwrapping_tsp_2014} and there is a desire for continuity in the notation between these papers.
